Palmer amaranth (*Amaranthus palmeri*) adaptation to US Midwest agroecosystems

### 2021-08-24 Frontiers/Frontiers.Rmd
as

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is one of the most troublesome weed species in the United States. Palmer amaranth is endemic in the Southern United States but its range is expanding northward. Palmer amaranth dispersal warrants studies assessing species adaptation into new geographies. A study was conducted to investigate morphology, flowering and sex dimorphism from cohorts of Palmer amaranth growing under corn, soybean, and bareground across five locations of United States Midwest. In general, results demonstrated that first cohort of Palmer amaranth, established in June, produced 42\% more biomass than plants from second cohort (established in July). At first cohort, Palmer amaranth produced 75.5 g plant\textsuperscript{-1} in bareground, 28.3 g plant\textsuperscript{-1} in soybean and 16.3 g plant\textsuperscript{-1} in corn, whereas Palmer amaranth produced 62.6, 6.3, and 1.4 g plant\textsuperscript{-1} at second cohort in bareground, soybean and corn, respectively. The plant height was more impacted by Palmer amaranth growing in corn. Palmer amaranth plants averaged 85.2 cm tall in the first cohort but 38.2 cm tall in the second cohort in corn. Moreover, Palmer amaranth flowering window shifted according to crop and cohort timings. Palmer amaranth growing in intense competition, such as under low light in corn, resulted in the longest flowering window. Also, Palmer amaranth sex dimorphism was slightly influenced by day of year, weight and height. The model estimated that probability of being a female plant increase as biomass and height increases. Our results showed the fast adaptation and plasticity of Palmer amaranth to grow and adapt to cropping systems. Palmer amaranth is likely to continue its expansion northward. Therefore, preventing plant dispersal into new habitat is the most effective management strategy. Reactive management to reduce Palmer amaranth impact on cropping systems should encompass diversity of tactics that minimize the species ability establish into cropping systems, including crop rotation (not only corn and soybean), early/late crop planting, row spacing, and cover crops. Our results also shed a light on sex dimorphism as a future Palmer amaranth ecological management.

@@

Palmer amaranth is a fast growing summer annual forb indigenous to the Sonoran Desert [@sauer1957]. The species would eventually emerge as a threat to US agriculture in the 1990s. Palmer amaranth weediness is likely a result of human-assisted selection in combination with plant biology. Farm mechanization, adoption of conservation agriculture (e.g., no-till), and reliance on herbicides for weed management are the main human-mediated selection of Palmer amaranth into cropping systems [@ward2013]. On the other hand, Palmer amaranth is a prolific seed producer with a C4 photosynthetic apparatus [@wang1992]. With a dioecy nature, Palmer amaranth male and female plants are obligate outcrosser species, increasing the chances of exchanging adaptive traits among plants [@jhala2021; @oliveira2018]. Also, Palmer amaranth small seed size (e.g, 1 mm) tend to thrive in no-tillage systems [@price2011], and spread across locations through farm equipment [@sauer1972], manure [@hartzler2016], animals [@farmer2017]. The dispersal capacity of Palmer amaranth makes the species one of the most successful cases of weed adaptation to current cropping systems.

@@

Palmer amaranth is already found in agronomic crops of South America [@larran2017; @kupper2017] and Southern Europe [@milani2021]. In the United States, Palmer amaranth is well established in the Cotton Belt [@garetson2019; @bagavathiannan2016] in the southern United States but its range is expanding northward. For example, herbicide resistant Palmer amaranth is widespread in Nebraska  [@oliveira2021a]. There are some reported cases of Palmer amaranth in Michigan [@kohrt2017] and Connecticut [@aulakh2021]. Also, it is estimated that Palmer amaranth can cause high damage to soybean fields in Illinois [@davis2015], which is concerning as soybean along with corn make most of US Midwest agronomic hectares. In Iowa, a study showed that Palmer amaranth is still not well adapted compared to waterhemp (*Amaranthus tuberculatus*) [@baker2021]. Invasion and successful eradication of Palmer amaranth is documented in Minnesota [@yu2021]. Palmer amaranth infestations have not been detected in Canada; however, Palmer amaranth seeds were detected in sweet potato slips in the country [@page2021]. Palmer amaranth is still not as well adapted and established to Northern as it is in the Southern North America. Therefore, Palmer amaranth range of expansion into new habitats is increasing. It seems fated the need to manage Palmer amaranth in agronomic crops throughout multiple environments in the near future. Strategies on Palmer amaranth management should encompass the agroecosystem level but not only attempts to eradicate the weed. Most tactics to manage Palmer amaranth are based on technology fixes [@scott2011], which are short-term (e.g., herbicide and/or tillage) rather than long-term weed management. Palmer amaranth management should be built on minimizing the species ability to adapt, grow and develop into agroecossystems.

@@

The experiment was conducted in 2018 and 2019 under field conditions at five locations: Arlington, WI (43°18'N, 89°29'W), Clay Center, NE (40.57'N, 9814'W), Grant, NE (('N, 'W)), Lincoln, NE ((41.16'N, 96.42'W)), and Macomb, IL (('N, 'W)).

@@

```{r echo=FALSE, warning=FALSE, message=FALSE}
options(knitr.kable.NA = '')
read_csv("../data/table.csv") %>% 
  janitor::row_to_names(row_number = 1) %>% 
  kable(booktabs = TRUE, caption = "Field study attributes ",
        col.names = c("Attributes",
                           "",
                           "Arlington, WI",
                           "Clay Center, NE",
                           "Grant, NE",
                      "Lincoln, NE", "Macomb, IL")) %>% 
#  pack_rows(
 #   index = c("Bareground" = 2, "Corn" = 4, "Soybean" = 8)) %>% 
  kable_styling(font_size = 10, 
                full_width = FALSE, 
                position = "center",
                latex_options = c("scale_down", "hold_position")) %>% 
  collapse_rows(columns = 1, latex_hline = "none", valign = "middle") %>%
  footnote(
           alphabet = c("glyphoste, 840 g ae ha", "S-metolachor, g ai ha", 
                        "S-metolachlor + trazine + mesotrione, + bicyclopyrone, g ai ha",
                        "saflufenacil + imazethapyr + pyroxasulfone, g ai ha"),
           footnote_as_chunk = TRUE)
```

@@

The field experimental units were three adjacent 9.1 m wide (12 rows at 76.2 cm row spacing) by 10.7 m long. The experimental design were arranged in factorial design with three crops, two transplanting times simulating two cohorts, repeated across five locations. Each experimental unit was planted with corn, soybean, or kept under bareground. The two transplanting timings were June 1 (first cohort) and July 1 (second cohort). Palmer amaranth seedlings (potting mix + two seedlings) were transplanted (6 cm deep and 8 cm wide). Forty-eight plants were equidistantly placed (0.76 m apart) between rows within each crop (Figure 2). After a week, one plant was eliminated and one was kept, resulting in 24 plants per experimental unit and transplanting time (Figure 2). When needed, Palmer amaranth plants were supplied with water during the first week after transplanting.

@@

The statistical analyses were performed using R statistical software version 4.0.1. Data analyses are stored at Zenodo [@oliveira2021b].

Analyses of Palmer amaranth height and biomass were performed with a linear mixed model using *lmer* function from "lme4" package [@bates2015]. Plant height and biomass were log transformed to meet model assumption of normality. In the model, crop (bareground, corn, soybean) was the fixed effect and year nested with location the random effects. Analysis of variance was performed with *anova* function from "car" package [@fox2018]. Marginal means and compact letter display were estimated with *emmeans* and *cld* from packages "emmeans" and "multcomp" [@hothorn2008].

The Palmer amaranth flowering timing was estimated as cumulative flowering across all location. Dataset was nested but id did not include Grant, NE. Palmer amaranth cumulative flowering estimation was determined using an asymmetrical three parameter log logistic Weibull model of the drc package [@ritz2015].

@@

Palmer amaranth sex dimorphism was fitted to a binary logistic regression [@bangdiwala2018]. Binary logistic regression is used for predicting binary classes, such as the probability of a plant being female in a dioecious species. Prior to the analysis, all missing values were removed from the dataset. Also, data from Grant was not used in this analysis due to the uniform plant harvesting at that location. The complete dataset was splitted into 80% train and 20% test data. The 80% train is used for the model training and the 20% test is used for checking how the model generalized on unseen dataset. With 80% dataset, a binary response variable, male (0) and female (1), was fitted to a generalized linear model (*glm* function) including day of year harvest, height, weight, crop and month as independent variables (without interaction). The model family was binomial with a logit function. The model fit was assessed through pseudo R-squared values (McFadden, Cox and Snell, Cragg and Uhler) and likelihood ratio using *nagelkerke* function ("rcompanion" package). The marginal effects computation was performed with Average Marginal Effects (AMEs) at every observed value of x and average across the results [@leeper2017] using *margins* function from "margins" package. The rest 20% dataset was predicted using *predict* function with a cutoff estimation for male or female using *performance* function. The model quality prediction from the classification algorithm was measured with precision (*precision* function), recall (*recall* function) and F1-score (*f_meas* function) using the "yardstick" package. The precision determines the accuracy of positive predictions (female plants), recall determines the fraction of positives that were correctly identified, and F1-score is a weighted harmonic mean of precision and recall with the best score of 1 and the worst score of 0. F1-score conveys the balance between the precision and the recall [@yacouby2020].

@@

The model fit was 0.23, 0.32, 0.40 using pseudo R-squared test from McFadden, Cox and Snell, and Cragg and Uhler, respectively. The likelihood ratio test showed a p-value of < 0.00. The average marginal effects showed that Palmer amaranth growing in corn resulted in 14.8% less females plants (Table 2). Moreover, increasing a unit doy  increases the probability of having a female plant by 0.4% (Table 2 and Figure4A). Similar trend is observed for weight as well as height, whereas the probability of being female increase by 0.2% (Figure 4B) and 0.1% (Figure 4C) when a unit of weight (g) and height (cm) increases, respectively.

@@

table_a %>% 
  kable(booktabs = TRUE, escape = TRUE, caption = "Average marginal means of sex dimorphism logistic model. Factor pararemter values (e.g. crop) is shown related to soybean.") %>% 
  kable_styling(font_size = 10, 
                full_width = FALSE, 
                position = "center",
                latex_options = c("hold_position")) %>% 
  collapse_rows(columns = 1, latex_hline = "none", valign = "middle") %>%
  footnote(
           alphabet = c("Average Marginal Effects.", "Standard Error."),
           footnote_as_chunk = TRUE)
```

@@

The model accuracy evaluation accuracy in the 20% test dataset was 0.62 with a cutoff value for female and male plants of 0.43. The model classification showed a precision of 0.64, recall of 0.66, and a F means score of 0.65. In addition, the area under the curve was 0.64.

@@

Our study showed that Palmer amaranth biomass, height, flowering pattern and sex varied within crops and cohort timings.  In general, first cohort of Palmer amaranth plants were heavier and taller when compared to the second cohort. At first cohort, resources (e.g., soil nutrients) and conditions (e.g., light) were more timely available for the species. High biomass and taller Palmer amaranth plants are likely a weed strategy to compete for light in between crop rows in absence of canopy. In such conditions, Palmer amaranth showed an extraordinary plasticity to adapt upon the agroecosystem. This is evident when comparing Palmer amaranth canopy shape, and its extended flowering pattern when growing into corn compared to soybean. The Palmer amaranth competition (e.g., light) strategy was to mimic the crop grow and development (Figure 6). These results suggests that Palmer amaranth can fast evolve life-history traits to adapt into cropping systems and cultural practices, which was also showed in a study varying nitrogen fertilization [@bravo2018]. Our results highlight the Palmer amaranth as a threat to field crops as breeding more competitive crop varieties is likely to select more competitive biotypes [@bravo2017].

@@

Palmer amaranth grow and development in second cohort was limited due to the crop competitive ability at advanced development stages. Palmer amaranth was transplanted when corn canopy was nearly closed, which reduced Palmer amaranth competitiveness. As a result, Palmer amaranth height and biomass was lower compared to its first cohort. Under crop canopy (e.g., second cohort), Palmer amaranth flowering window was near to similar in corn and soybean. Palmer amaranth growing without crop competition produced the highest amounts of biomass and less extended flowering window. The Palmer amaranth strategy in bareground was to invest biomass in growing plant width and height. Nonetheless, Palmer amaranth produced 21% less biomass in second cohort compared to first cohort timing. In a bareground study, early emerged Palmer amaranth without competition was 50% taller than late emerged plants [@webster2015]. These results suggest that crop competition is not the only factor limiting late Palmer amaranth establishment. The limited growth of Palmer amaranth at second cohort is likely a reduced plant response to day length, light availability and thermal units (e.g, growing degree days). It is hypothesize that reduced day length contributed to smaller plants at second cohort as well as shorter flowering period. A study in North Carolina and Illinois predicted that less than 10% Palmer amaranth seedlings emergence occurred after June [@piskackova2021]. In addition, Palmer amaranth negative impact on soybean [@korres2020] and cotton [@webster2015] yields was higher when plants were established near to crop planting.

Seed production was not evaluated due to plant harvest at initiation of flowering. Nonetheless, it is well documented a strong positive correlation between Palmer amaranth biomass and seed production [@schwartz2016; @spaunhorst2018]. I our study, Palmer amaranth growing at first cohort accumulated an overall 42% more biomass when compared to second cohort. Therefore, Palmer amaranth plants growing in the second cohort is likely to produce less seeds regardless the cropping system. Our observation is consistent with the findings that first Palmer amaranth cohort produced 50% more seeds per plant than Palmer amaranth plants established six weeks later in bareground [@webster2015]. Still, seed production at second cohort is likely to replenish the soil seedbank. Seed production and deposition in the seedbank is also a key factor for species perpetuation [@menges1987]. Palmer amaranth can produce hundred thousands seeds per plant [@schwartz2016; @keeley1987], and stay viable buried in the seedbank for at least 36 months [@sosnoskie2013]. Therefore, preventing Palmer amaranth seed production or/and seed migration to its non-native habitat is an essential strategy to minimize weed impact into agroecossystem [@davis2015].

An ecological approach to reduce seed production in Palmer amaranth is understanding its flowering window. Our study suggests that Palmer amaranth flowering was slightly influenced by cropping systems and cohort timings. Palmer amaranth growing in bareground and corn resulted in the overall shortest and longest flowering window, respectively. When growing in soybean, Palmer amaranth flowering window was similar to bareground at first cohort but similar to corn at second cohort timing. Plant flowering initiation is complex and depends on biological and ecological factors [@lang1965]. We hypothesize that when growing in high competition (e.g., second cohort) and/or under low light/short day length, Palmer amaranth plants tend to initiate flowering early, as well as having an extended flowering window. Early flower initiation is plant strategy when growing in stress conditions. For example, when growing under water stress, early flowering in Palmer amaranth resulted in a mismatch between female and male plants by seven days [@mesgaran2021]. A mismatch in Palmer amaranth male and female flowering period can minimize plant outcross, and thus reduce plant seed production and exchange of resistant alleles [@jhala2021]. Sex dimorphism manipulation is considered a potential ecological pest control [@mcfarlane2018; @schliekelman2005].

The mechanisms of sex-determination in plant species is intriguing and arouse the curiosity of many scientists, including Charles Darwin [@darwin1888]. In our study, the sex dimorphism model performance was decent (AIC 0.64) considering the biology of plant flowering. A 1:1 male and female sex ratio is a general evolutionary stable strategy for plant species perpetuation [@fisher1930]. However, a slight deviation from 1:1 sex ratio might occur in some dioecious species. For example, the dioecious *Halophila stipulacea* is a female-biased plant in its native habitat, but the naturalized *H. stipulacea* have a 1:1 ratio [@nguyen2018]. Naturalized of *H. stipulacea* reduced female-male ratio to expand into its non-native habitat [@nguyen2018]. Also, biotic and/or abiotic stress can influence plant sex determination. Palmer amaranth  male-to-female ratio was greater under high plant densities [@korres2017a] and after herbicide application [@rumpa2019]. Our model estimated that late flowering, heavier and taller Palmer amaranth plants deviated from 1:1 ratio in favor to female plants. It was reported that female Palmer amaranth plants invested more in height, stem and biomass while male invested more in leaf area and leaf dry weight under nutrient deficiency [@korres2017]. Our model also estimated more female plants in soybean and bareground compared to corn, which might linked to plant competition strategy in each agroecossystem. Our results showed the influence of life-history and ecological traits on sexual dimorphism in Palmer amaranth. Sexual dimorphism is documented in other dioecious species [@barrett2013]. For example, stronger female plant competition and greater male tolerance to herbivory was reported in *Spinacia oleracea* [@perez-llorca2019]. Research on candidate genes for sex determination in *Amaranthus* species are currently underway but it is far to complete [@montgomery2021; @montgomery2019]. Further studies are also needed to understand the ecological basis of Palmer amaranth flowering, including the plant behavior under climate change.

Our study demonstrated the Palmer amaranth plasticity to grow and develop into cropping-systems. Is likely that Palmer amaranth range will continue to expanding into new geographies. Therefore, preventive management is a priority to minimizing Palmer amaranth dispersal. Reactive management should focus on early-season management programs, which would have a large negative effect on Palmer amaranth growth and development. Tactics that promote early-season crop advantage against Palmer amaranth, including early crop planting, crop rotation, plant width, preemergence applied herbicide, and crop residue (e.g. cover crops) would minimize the negative impact of Palmer amaranth in agroecossystems. The aggressiveness and differential Palmer amaranth adaptation to agroecosystem is striking and require national efforts to minimize the species impact on economy and sustainability.

@@

```{r, Figure-2, ref.label = "fig2", echo = FALSE, message = FALSE, fig.height=3, fig.width=6, fig.align='center', fig.cap='Palmer amaranth adaptation study layout of a plant cohort timing. Twenty-four Palmer amaranth plants were place 76.2 cm apart in each experimental unit', out.width = "150mm", out.height = "70mm"}
# You can also refer to code chunks from above to place figures at the bottom.
```

>>>>

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is one of the most troublesome weed species in the United States. Palmer amaranth is endemic to Southern United States but its range is expanding northward. Palmer amaranth dispersal warrants studies assessing species adaptation into new geographies. A study was conducted to investigate morphology, flowering and sex dimorphism from cohorts of Palmer amaranth growing under corn, soybean, and bareground across five locations of US Midwest. In general, results demonstrated that first cohort of Palmer amaranth, established in June, produced 42\% more biomass than plants from second cohort (established in July). At first cohort, Palmer amaranth produced 75.5 g plant\textsuperscript{-1} in bareground, 28.3 g plant\textsuperscript{-1} in soybean and 16.3 g plant\textsuperscript{-1} in corn, whereas Palmer amaranth produced 62.6, 6.3, and 1.4 g plant\textsuperscript{-1} at second cohort in bareground, soybean and corn, respectively. The plant height was more impacted by Palmer amaranth growing in corn. Palmer amaranth plants averaged 85.2 cm tall in the first cohort but 38.2 cm tall in the second cohort in corn. Moreover, Palmer amaranth flowering window shifted according to crop and cohort timings. Palmer amaranth growing in intense competition, such as under low light in corn, resulted in the longest flowering window. Also, Palmer amaranth sex dimorphism was slightly influenced by day of year, weight and height. The model estimated that probability of being a female plant increase as biomass and height increases. Our results showed the fast adaptation and plasticity of Palmer amaranth to grow and adapt to cropping systems from United States Midwest. Palmer amaranth is likely to continue its expansion northward. Therefore, preventing plant dispersal into new habitats is the most effective management strategy. Reactive management to reduce Palmer amaranth impact on cropping systems should encompass diversity of tactics that minimize the species ability establish into cropping systems, including crop rotation (not only corn and soybean), early/late crop planting, row spacing, and cover crops. Our results also shed a light on sex dimorphism as a future Palmer amaranth ecological management.

@@

Palmer amaranth is a fast growing summer annual forb indigenous to the Sonoran Desert [@sauer1957]. The species would eventually emerge as a threat to US agriculture in the 1990s. Palmer amaranth weediness is likely a result of human-assisted selection in combination with plant biology. Farm mechanization, adoption of conservation agriculture (e.g., no-till), and intensive use of herbicides for weed management are the main human-mediated selection of Palmer amaranth into cropping systems [@ward2013]. On the other hand, Palmer amaranth is a prolific seed producer with a C4 photosynthetic apparatus [@wang1992]. With a dioecy nature, Palmer amaranth male and female plants are obligate outcrosser species, increasing the chances of exchanging adaptive traits among plants [@jhala2021; @oliveira2018]. Also, Palmer amaranth small seed size (e.g, 1 mm) tend to thrive in no-tillage systems [@price2011], and spread across locations through farm equipment [@sauer1972], manure [@hartzler2016], animals [@farmer2017]. The dispersal capacity of Palmer amaranth makes the species one of the most successful cases of weed adaptation to current cropping systems.

@@

Palmer amaranth is already found in agronomic crops of South America [@larran2017; @kupper2017] and Southern Europe [@milani2021]. In the United States, Palmer amaranth is well established in the Cotton Belt [@garetson2019; @bagavathiannan2016] in the southern United States but its range is expanding northward. For example, herbicide resistant Palmer amaranth is widespread in Nebraska  [@oliveira2021a]. There are some reported cases of Palmer amaranth in Michigan [@kohrt2017] and Connecticut [@aulakh2021]. Also, it is estimated that Palmer amaranth can cause high damage to soybean fields in Illinois [@davis2015], which is concerning as soybean along with corn make most of US Midwest agronomic hectares. In Iowa, a study showed that Palmer amaranth is still not well adapted compared to waterhemp (*Amaranthus tuberculatus*) [@baker2021].  Waterhemp, an *Amaranthus* species, is the most troublesome species in that geography [@tranel2011]. Nonetheless, invasion and successful eradication of Palmer amaranth is documented in Minnesota [@yu2021]. Palmer amaranth infestations have not been detected in Canada; however, Palmer amaranth seeds were detected in sweet potato slips in the country [@page2021]. Palmer amaranth is still not as well adapted and established to Northern as it is in the Southern North America. Therefore, Palmer amaranth range of expansion into new habitats can increase. It seems certain the need to manage Palmer amaranth in agronomic crops throughout novel environments in the near future. Strategies on Palmer amaranth management should encompass the agroecosystem level but not only attempts to eradicate the weed. Most tactics to manage Palmer amaranth are based on technology fixes [@scott2011], which are short-term (e.g., herbicide and/or tillage) rather than long-term weed management. Palmer amaranth management should be built on minimizing the species ability to adapt, grow and develop into agroecossystems.

@@

The experiment was conducted in 2018 and 2019 under field conditions at five locations: Arlington, WI (43°18'N, 89°29'W), Clay Center, NE (40.57'N, 9814'W), Grant, NE ('N, 'W), Lincoln, NE (41.16'N, 96.42'W), and Macomb, IL ('N, 'W).

@@

```{r echo=FALSE, warning=FALSE, message=FALSE}
options(knitr.kable.NA = '')
read_csv("../data/table.csv") %>% 
  janitor::row_to_names(row_number = 1) %>% 
  kable(booktabs = TRUE, caption = "Field study attributes ",
        col.names = c("Attributes",
                           "",
                           "Arlington, WI",
                           "Clay Center, NE",
                           "Grant, NE",
                      "Lincoln, NE", "Macomb, IL")) %>% 
#  pack_rows(
 #   index = c("Bareground" = 2, "Corn" = 4, "Soybean" = 8)) %>% 
  kable_styling(font_size = 10, 
                full_width = FALSE, 
                position = "center",
                latex_options = c("scale_down", "hold_position")) %>% 
  collapse_rows(columns = 1, latex_hline = "none", valign = "middle") %>%
  footnote(
           alphabet = c("glyphoste, 840 g ae ha", "S-metolachor, 1324 g ai ha;", 
                        "S-metolachlor + trazine + mesotrione, + bicyclopyrone, 2409 g ai ha;",
                        "saflufenacil + imazethapyr + pyroxasulfone, 215 g ai ha"),
           footnote_as_chunk = TRUE)
```

@@

The field experimental units were three adjacent 9.1 m wide (12 rows at 76.2 cm row spacing) by 10.7 m long. The experimental design were arranged in factorial design with three crops, two transplanting times simulating two cohorts, repeated across five locations. Each experimental unit was planted with corn, soybean, or kept under bareground. The two transplanting timings were June 1 (first cohort) and July 1 (second cohort). Palmer amaranth seedlings (potting mix + two seedlings) were transplanted (6 cm deep and 8 cm wide). Forty-eight plants were equidistantly placed (0.76 m apart) between rows within each crop. After a week, one plant was eliminated and one was kept, resulting in 24 plants per experimental unit and transplanting time (Figure 2). When needed, Palmer amaranth plants were supplied with water during the first week after transplanting.

@@

The statistical analyses were performed using R statistical software version 4.0.1 [@rcoreteam2021]. Data analyses are stored at Zenodo [@oliveira2021b].

Analyses of Palmer amaranth height and biomass were performed with a linear mixed model using *lmer* function from "lme4" package [@bates2015]. Plant height and biomass were log transformed to meet model assumption of normality. In the model, crop (bareground, corn, soybean) and cohort time (first and second) were the fixed effects and year nested with location the random effects. Analysis of variance was performed with *anova* function from "car" package [@fox2018]. Marginal means and compact letter display were estimated with *emmeans* and *cld* from packages "emmeans" [@lenth2021] and "multcomp" [@hothorn2008], respectively.

The Palmer amaranth flowering timing was estimated as cumulative flowering across all location. Dataset was nested but it did not include Grant, NE. Palmer amaranth cumulative flowering estimation was determined using an asymmetrical three parameter log logistic Weibull model of the drc package [@ritz2015].

@@

Palmer amaranth sex dimorphism was fitted to a binary logistic regression [@bangdiwala2018]. Binary logistic regression is used for predicting binary classes, such as the probability of a plant being female in a dioecious species. Prior to the analysis, all missing values were removed from the dataset. Also, data from Grant was not used in this analysis due to the uniform plant harvesting at that location. The complete dataset was splitted into 80% train and 20% test data. The 80% train is used for the model training and the 20% test is used for checking model performance on unseen dataset. With 80% dataset, a binary response variable, male (0) and female (1), was fitted to a generalized linear model (base R *glm* function) including day of year harvest, height, weight, crop and month as independent variables (without interaction). The model family was binomial with a logit function. The model fit was assessed through pseudo R-squared values (McFadden, Cox and Snell, Cragg and Uhler) and likelihood ratio using *nagelkerke* function from "rcompanion" package [@mangiafico2021]. The marginal effects computation was performed with Average Marginal Effects (AMEs) at every observed value of x and average across the results [@leeper2017] using *margins* function from "margins" package [@leeper2021]. The rest 20% dataset was predicted using *predict* function with a cutoff estimation for male or female using *performance* function. The model quality prediction from the classification algorithm was measured with precision (*precision* function), recall (*recall* function) and F1-score (*f_meas* function) using the "yardstick" package [@kuhn2021]. The precision determines the accuracy of positive predictions (female plants), recall determines the fraction of positives that were correctly identified, and F1-score is a weighted harmonic mean of precision and recall with the best score of 1 and the worst score of 0. F1-score conveys the balance between the precision and the recall [@yacouby2020].

@@

The model fit was 0.23, 0.32, 0.40 using pseudo R-squared test from McFadden, Cox and Snell, and Cragg and Uhler, respectively. The likelihood ratio test showed a p-value of < 0.00. The average marginal effects showed that Palmer amaranth growing in corn resulted in 14.8% less females plants (Table 2). Moreover, increasing a unit doy increases the probability of having a female plant by 0.4% (Table 2 and Figure4A). Similar trend is observed for weight (Figure 4B) as well as height (Figure 4C), whereas the probability of being female increase by 0.2% and 0.1% when a unit of weight (g) and height (cm) increases, respectively.

@@

table_a %>% 
  kable(booktabs = TRUE, escape = TRUE, caption = "Average marginal effects of Palmer amaranth sex dimorphism logistic model. Factor pararemter values (crop and bareground) is shown compared to soybean.") %>% 
  kable_styling(font_size = 10, 
                full_width = FALSE, 
                position = "center",
                latex_options = c("hold_position")) %>% 
  collapse_rows(columns = 1, latex_hline = "none", valign = "middle") %>%
  footnote(
           alphabet = c("Average Marginal Effects.", "Standard Error."),
           footnote_as_chunk = TRUE)
```

@@

The model accuracy evaluation accuracy in the 20% test dataset was 0.62 with a cutoff value for female and male plants of 0.43. The model classification showed a precision of 0.64, recall of 0.66, and a F1-score of 0.65. In addition, the area under the curve was 0.64.

@@

Our study showed that Palmer amaranth biomass, height, flowering window and sex varied within crops and cohort timings.  In general, first cohort of Palmer amaranth plants were heavier and taller when compared to the second cohort. At first cohort, resources (e.g., soil nutrients) and conditions (e.g., light) were more timely available for the species. High biomass and taller Palmer amaranth plants are likely a weed strategy to compete for light in between crop rows in absence of canopy. In such conditions, Palmer amaranth showed an extraordinary plasticity to adapt upon the agroecosystem. This is evident when comparing Palmer amaranth canopy shape. The Palmer amaranth competition (e.g., light) strategy was to mimic the crop grow and development (Figure 6). These results suggests that Palmer amaranth can fast evolve life-history traits to adapt into cropping systems and cultural practices, which was also showed in a study varying nitrogen fertilization [@bravo2018]. Our results highlight the Palmer amaranth as a threat to field crops as breeding more competitive crop varieties is likely to select more competitive weed biotypes [@bravo2017].

@@

Palmer amaranth grow and development in second cohort was limited due to the crop competitive ability at advanced development stages. Plants were transplanted at greater crop height, which reduced Palmer amaranth competitiveness. As a result, Palmer amaranth height and biomass was lower compared to its first cohort. Palmer amaranth growing without crop competition produced the highest amounts of biomass. The Palmer amaranth strategy in bareground was to invest biomass in growing plant width and height. Nonetheless, Palmer amaranth produced 17% less biomass in second cohort compared to first cohort timing. In a bareground study, early emerged Palmer amaranth without competition was 50% taller than late emerged plants [@webster2015]. These results suggest that crop competition is not the only factor limiting late Palmer amaranth establishment. The limited growth of Palmer amaranth at second cohort is likely a reduced plant response to day length, light availability and thermal units (e.g, growing degree days). The *Amaranthus* species are sensitive to photoperiod [@wu2014]. It is hypothesize that reduced day length contributed to smaller plants at second cohort regardless the crop. A study in North Carolina and Illinois predicted that less than 10% Palmer amaranth seedlings emergence occurred after June [@piskackova2021]. In addition, Palmer amaranth negative impact on soybean [@korres2020] and cotton [@webster2015] yields was higher when plants were established near to crop planting.

Seed production was not evaluated due to plant harvest at initiation of flowering. Nonetheless, it is well documented a strong positive correlation between Palmer amaranth biomass and seed production [@schwartz2016; @spaunhorst2018]. I our study, plants growing at first cohort accumulated an overall 42% more biomass when compared to second cohort. Therefore, Palmer amaranth plants growing in the second cohort is likely to produce less seeds regardless the crop. Our observation is consistent with the findings that first Palmer amaranth cohort produced 50% more seeds per plant than Palmer amaranth plants established six weeks later in bareground [@webster2015]. Still, seed production at second cohort is likely to replenish the soil seedbank. Seed production and deposition in the seedbank is also a key factor for species perpetuation [@menges1987]. Palmer amaranth can produce hundred thousands seeds per plant [@schwartz2016; @keeley1987], and stay viable buried in the seedbank for at least 36 months [@sosnoskie2013]. Therefore, preventing Palmer amaranth seed production or/and seed migration to its non-native habitat is an essential strategy to minimize the species impact into agroecossystem [@davis2015].

An ecological approach to reduce seed production in Palmer amaranth is understanding its flowering window. Our study suggests that Palmer amaranth flowering was slightly influenced by crops and cohort timings. Palmer amaranth growing in bareground and corn resulted in the overall shortest and longest flowering window, respectively. The shortest flowering window of second cohorts of Palmer amaranth growing in bareground highlighted the impact of photoperiod on flowering. When growing in soybean, Palmer amaranth flowering window was similar to bareground at first cohort but similar to corn at second cohort timing. Plant flowering initiation is complex and depends on biological and ecological factors [@lang1965]. We hypothesize that when growing in high competition (e.g., second cohort), Palmer amaranth plants tend to initiate flowering early, as well as having an extended flowering window. A study has shown that non-native Palmer amaranth initiated flowering two weeks prior to the native waterhemp in Iowa [@baker2021]. Early flower initiation could be also a plant strategy when growing in stress conditions. For example, when growing under water stress, early flowering in Palmer amaranth resulted in a mismatch between female and male plants by seven days [@mesgaran2021]. A mismatch in Palmer amaranth male and female flowering period can minimize plant outcross, reducing plant seed production and/or exchange of resistant alleles [@jhala2021]. Sex dimorphism manipulation is considered a potential ecological pest control [@mcfarlane2018; @schliekelman2005].

The mechanisms of sex-determination in plant species is intriguing and arouse the curiosity of many scientists, including Charles Darwin [@darwin1888]. In our study, the sex dimorphism model performance was decent (AIC 0.64) considering the biology of plant flowering. A 1:1 male and female sex ratio is a general evolutionary stable strategy for plant species perpetuation [@fisher1930]. However, a slight deviation from 1:1 sex ratio might occur in some dioecious species. For example, the dioecious *Halophila stipulacea* is a female-biased plant in its native habitat, but the naturalized *H. stipulacea* have a 1:1 ratio [@nguyen2018]. Naturalized of *H. stipulacea* reduced female-male ratio to expand into its non-native habitat [@nguyen2018]. Also, biotic and/or abiotic stress can influence plant sex determination. Palmer amaranth  male-to-female ratio was greater under high plant densities [@korres2017a] and after herbicide application [@rumpa2019]. Our model estimated that late flowering, heavier and taller Palmer amaranth plants deviated from 1:1 ratio in favor to female plants. It was reported that female Palmer amaranth plants invested more in height, stem and biomass while male invested more in leaf area and leaf dry weight under nutrient deficiency [@korres2017]. Our model also estimated more female plants in soybean and bareground compared to corn, which might linked to plant competition strategy in each crop. Our results showed the influence of life-history and ecological traits on sexual dimorphism in Palmer amaranth. Sexual dimorphism is documented in other dioecious species [@barrett2013]. For example, stronger female plant competition and greater male tolerance to herbivory was reported in *Spinacia oleracea* [@perez-llorca2019]. Research on candidate genes for sex determination in *Amaranthus* species are currently underway but it is far to complete [@montgomery2021; @montgomery2019]. Further studies are also needed to understand the ecological basis of Palmer amaranth flowering, including the plant behavior under climate change.

Our study demonstrated the Palmer amaranth plasticity to grow and develop into arable land of US Midwest. Is likely that Palmer amaranth range will continue to expanding into new geographies. The migration of Palmer amaranth into the US Midwest will reshape the landscape as waterhemp and Palmer amaranth will share the same habitat. The presence of Palmer amaranth and waterhemp will increase the weed management complexity. Therefore, preventive management is a priority to minimizing Palmer amaranth dispersal. Reactive management should focus on early-season management programs, which would have a large negative effect on Palmer amaranth growth and development. Long-term tactics that promote early-season crop advantage against Palmer amaranth, including diversity of crops in rotation, early/late crop planting, plant width, and crop residue (e.g. cover crops) would minimize the negative impact of Palmer amaranth to cropping systems. The aggressiveness and differential Palmer amaranth adaptation to agroecosystem is striking and require national efforts to minimize the species impact on sustainability and economy.

@@

```{r, Figure-2, ref.label = "fig2", echo = FALSE, message = FALSE, fig.height=3, fig.width=6, fig.align='center', fig.cap='Palmer amaranth adaptation study layout of a plant cohort timing in bareground, corn, and soybean. Twenty-four Palmer amaranth plants were place 76.2 cm apart in each experimental unit', out.width = "150mm", out.height = "70mm"}
# You can also refer to code chunks from above to place figures at the bottom.
```
